---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most?
**A1.** There are 26 departments with classes with climate in the name. Atmospheric and Oceanic Science

```{r}
umd_courses |>
  filter(str_detect(title, "Climate")) |> filter(seats >= 1) |>  
  group_by(department) |> 
  summarize(total = sum(count_classes = n())) |> 
  arrange(desc(total))
```

**Q2.** Of the departments listed in A2, which one is the most interesting to you, and why?
**A2.** TLPL is the most interesting to me because I have taken a TLPL class before and I really enjoyed it.I would probably want to investigate AOSC more because it has the highest total. 


**Q3** Which departments have the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? How are the top five departments different in each result?

**A3.** Business and Management, School of Music, English, Information Studies, and Journalism have the most courses in fall 2023. Business and Management, School of Music, and Information Studies have at least one syllabus posted. The total changes when adding the syllabus_count filter.

```{r}
courses_2023 <- umd_courses |> 
  filter(term > 202300) |> 
  group_by(department) |> 
  summarize(total = sum(count_classes = n())) |> 
  arrange(desc(total))

courses_2023 <- umd_courses |> 
  filter(term > 202300) |> filter(syllabus_count >= 1) |> 
  group_by(department) |> 
  summarize(total = sum(count_classes = n())) |> 
  arrange(desc(total))

head(courses_2023)
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as the same dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time.
**A4.** From May 2021 to June 2021, Moore's spending increased drastically. From June 2021 to April 2022, his spending increased pretty steadily. Then, he spend less in May 2022 than he did thr month prior. In June and July 2022, he spend a lot more than ever before. His spending then decreased in August 2022 and September 2022. His spending then increased by 6x more from September 2022 to October 2022, and then decreased again in November 2022.

```{r}
maryland_expenses <- read_csv("data/maryland_expenses.csv")

maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))

maryland_expenses |>
  mutate(month = floor_date(expenditure_date, "month")) |>
  filter(str_detect(committee_name, "Moore  Wes For Maryland")) |> 
  group_by(month) |>
  summarise(total_amount = sum(amount)) |>
  arrange(month)
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for the others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).
**A5** The results are confusing, as it is extremely unlikely that Wes Moore spent no money outside of Maryland. Although the code is showing what Q5 is asking for, the data does not align with what is realistic.

```{r}
maryland_expenses_with_state <- maryland_expenses |>
  mutate(
    spending_location = case_when(
        str_detect(address, " Maryland ") ~ "In-State",
        .default = "Out Of State"
      )
  )

maryland_expenses_with_state |>
  group_by(spending_location) |>
  summarize(total = sum(amount)) |>
  arrange(desc(total))
```


**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.
**A6.** How many courses are about data? Answer- 1066 courses

```{r}
data_courses <- umd_courses |>
  filter(str_detect(title, "Data")) |>  
  group_by(department) |> 
  summarize(total = sum(count_classes = n())) |> 
  arrange(desc(total))

data_courses |> summarize(grand_total = sum(total))
```
